---
title: "545_Quant_Week_3_DataPrep"
author: "Hye Min Park"
date: "1/24/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package, include=FALSE}
library(data.table)
library(plyr)

```

## about my data

#Weather variables
source: cal-adapt <https://cal-adapt.org/tools/extreme-heat/>

#Extreme Heat Days & Warm Nights
Number of Extreme Heat Days per Year when daily maximum temperature is above 103.9 °F 
Number of Warm Nights per Year when daily minimum temperature is above 67.2 °F

Datasets created from LOCA downscaled CMIP5 climate projections for Cal-Adapt tools. These currently include the modeled annual variability envelope (maximum and minimum from range of annual average values from all 32 GCMs); precalculated data tables of extreme heat counts for California counties for 4 priority models and 2 scenarios (RCP 4.5 and RCP 8.5)

RCP 4.5 is described by the Intergovernmental Panel on Climate Change (IPCC) as a moderate scenario in which emissions peak around 2040 and then decline. RCP 8.5 is the highest baseline emissions scenario in which emissions continue to rise throughout the twenty-first century.


#1. Historical CDD data


```{r read data}
setwd("C:/Users/hmpark/PATH/Git/hm/Quantitative-Analysis-Practicum-2024/student_work/HyeMin")

#load historical cooling degree days data
cdd_avg_his = read.csv("LACountyTemp/cdd_year_ens32avg_historical.csv")
setDT(cdd_avg_his) #convert data frame to data table
cdd_avg_his[, year := as.numeric(substr(time, 1, 4))] #extract year from time and convert chr to numeric
cdd_avg_his[,time:=NULL] #delete the time column
cdd_avg_his[, .(cdd_avg_hist=cdd_year_ens32avg_historical , year)] # change column name
colnames(cdd_avg_his)[1] <- "cdd_avg_hist"

head(cdd_avg_his)
```


#2. Future CDD data

```{r future data}
#repeat this for other data sets

cdd_avg_rcp45 = read.csv("LACountyTemp/cdd_year_ens32avg_rcp45.csv")
cdd_avg_rcp85 = read.csv("LACountyTemp/cdd_year_ens32avg_rcp85.csv")
setDT(cdd_avg_rcp45)
setDT(cdd_avg_rcp85)

cdd_avg_rcp45[, year := as.numeric(substr(time, 1, 4))] #extract year from time and convert chr to numeric
cdd_avg_rcp45[,time:=NULL] #delete the time column

cdd_avg_rcp45[, .(cdd_avg_rcp45 = cdd_year_ens32avg_rcp45 , year)] # change column name


cdd_avg_rcp85[, year := as.numeric(substr(time, 1, 4))] #extract year from time and convert chr to numeric
cdd_avg_rcp85[,time:=NULL] #delete the time column

cdd_avg_rcp85[, .(cdd_avg_rcp85=cdd_year_ens32avg_rcp85 , year)] # change column name

#joing rcp45 and 85
cdd_future <- merge(cdd_avg_rcp45,cdd_avg_rcp85, by="year")
colnames(cdd_future)[2] <- "cdd_avg_rcp45"  
colnames(cdd_future)[3] <- "cdd_avg_rcp85"  
head(cdd_future)

```
#load electricity load data

```{r load data}
load_w = read.csv("ElectricityByCounty.csv")
setDT(load_w)
load_w[,Sector:=NULL] #delete the column
load_w[,Total Usage:=NULL]
load <- melt(load_w, variable.name="year", value.name = "consumption") #wide to long
load[,year:=as.numeric(gsub("X","",year))] #remove X prefix and convert it to numeric
head(load)

```

Sum across the counties

```{r load data}
CA_load <- load[, sum(.SD$consumption), by="year"]
colnames(CA_load)[2] <- "total_load" 
head(CA_load)
```

#Merge historical cdd with load
```{r merge historical data}
historical <- merge(CA_load,cdd_avg_his, by="year")
plot(x=historical$cdd_avg_his,y=historical$CA_load)
```

